GEE models

### 2016-07-27 doc/manuscript.Rmd
Updated R code link in manuscript
Updated manuscript to fit with Diabetologia reqs

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
#csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
csl: /home/luke/Documents/bibliography/styles/vancouver-brackets-no-et-al.csl
output: fost::diabetes_care_manuscript
---

@@

<!-- 47 characters + spaces -->
Running title: NEFA and the pathogenesis of diabetes

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,3,6), PhD

<!-- During time of study. -->
Affiliation:

@@

Word count: 3620 / 4000

@@

<!-- Reference limit 39 / 40 -->

@@

Objective: Our aim was to determine the longitudinal associations of individual
non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and
beta-cell function over 6 years in a cohort of individuals who are at-risk for
diabetes.

Research Design and Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and completed an OGTT at 3 time points over
6 years. Outcome variables were calculated using the OGTT values. IS was
assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function was
assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, baseline age, ALT, and family
history of diabetes. NEFA were analyzed as both concentrations and proportions
(mol%) of the total fraction.

@@

@@

```{r caption_setup}
## Include captions below using `captioner` package
fig <- captioner(prefix = 'Figure')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner(prefix = 'Supplemental Figure S')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner(prefix = 'Supplemental Table S')
cite_st <- pryr::partial(stab, display = 'cite')
# usage: cite_st('basicChar')

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC, ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')

@@

sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%. See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot.')

@@

sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.')
sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA). Discriminatory ability is evident from the amount of separation between the two groups. In this case, PLS-DA was 82% accurate at correctly classifying groups. See the Supplemental Methods for a description of PLS-DA and for interpreting this plot.')
sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis. See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).  See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status. See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))

@@

# Background

@@

# Research Design and Methods

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. Therefore, to confirm the results of the GEE
analyses in a multivariate environment (all NEFA variables in a single model),
partial least squares discriminant analysis (PLS-DA) was used to identify the
patterns of NEFA composition against beta-cell function (using ISSI-2, which had
the majority of the associations in the GEE models) and for those
who converted to or maintained dysglycemia status (either IFG, IGT, or DM) over
the 6 years. Latent class mixed models (LCMM) were used to extract the underlying
trajectories of beta-cell function (ISSI-2) over the 6 years. For more detailed
explanation of GEE, PLS-DA, and LCMM, please see the Supplemental Methods.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE <!--[@Hoejsgaard2006a]-->, caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{add link before
submission}}. Results were considered statistically significance at p<0.05,
after adjusting for multiple testing using the Benjamini-Hochberg False
Discovery Rate [@Benjamini1995a].

@@

```{r inlineResults}
percent.nefa <- calculate_percent_nefa_contribution(ds) 
misclass.n.percent.bcf <- calculate_plsda_misclass(plsda_results)
misclass.n.percent.dysgly <- calculate_plsda_misclass(plsda_dysgly)
percent.change.outcomes <- calculate_outcomes_pct_change(ds)
lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
dm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertDM')
predm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertPreDM')
sex.npct <- calculate_factor_npercent(ds0yr$Sex, 'Female')
ethn.npct <- calculate_factor_npercent(ds0yr$Ethnicity, 'European')
fhd.npct <- calculate_factor_npercent(ds0yr$FamHistDiab, 'Yes')
est.gee <- calculate_gee_magnitude(gee_results)
```

@@

<!--
A limitation of these methods, excluding LCMM,
is that they can not be adjusted for covariates.
-->

In conclusion, we found that total NEFA was a strong predictor for lower
beta-cell function over 6 years, irrespective of the specific composition of the
NEFA fraction. While future studies are needed to confirm these findings, our
results reinforce the importance of continuing to investigate the role of
circulating NEFA concentration on the natural history of diabetes.

## Acknowledgements

The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
in Family Medicine, University of Western Ontario, London, Canada for their 
expert technical assistance and dedication in their work for PROMISE. 
The authors had the following responsibility: LWJ conducted research, analyzed
data, and wrote the paper; RR, ZL, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content. All authors read and approved the final manuscript. AJH is
the guarantor of this work and, as such, had full access to all the data in the
study and takes responsibility for the integrity of the data and the accuracy of
the data analysis.

This study was supported by grants from the Canadian Diabetes Association (CDA),
the Canadian Institutes for Health Research (CIHR), and the University of
Toronto Banting and Best Diabetes Centre (BBDC); LWJ is supported by a CDA
Doctoral Student Research Award; RR is supported by a Heart and Stroke
Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
National Diabetes Management and the Ian McWhinney Chair of Family Medicine
Studies at the University of Western Ontario; RBP holds a Tier II Canada
Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
Chair in Diabetes Epidemiology. The authors report no potential conflicts of
interest relevant to this study.

Portions of the results in the present analysis have been presented in
abstract form at the following conferences: International Symposium on Diabetes
& Nutrition (2015); Biennial Congress of the International Society for the Study
of Fatty Acids and Lipids (2014); International Diabetes Federation World
Diabetes Congress (2015); American Diabetes Association Scientific Sessions
(2014, 2015).

# Figures

@@

Note: Values are in median (IQR), mean (SD), and n (%). The proportion of ethnic
and sex groups did not change over the 6 years.

# Supplemental Material

## Supplemental Methods

@@

## Supplemental Tables and Figures

@@

```{r onlineSupplementalTableS1}
prep_qic <- gee_df %>%
    dplyr::select(
        SID, VN, lISSI2, lISI, TotalNE, Sex, Ethnicity, MET, ALT, BaseAge,
        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
        )

@@

```{r onlineSupplementalTableS2}
table_distribution(ds, stab_dist)
```

```{r onlineSupplementalFigureS2, fig.cap=sfig_corr}
plot_heatmap(ds)
```

```{r onlineSupplementalFigureS3, fig.cap=sfig_gee}
plot_gee_results(gee_results_unadj)
```

```{r onlineSupplementalTableS3_4}
table_gee(gee_results, stab_gee)
table_gee(gee_results_unadj, stab_gee_unadj)
```

```{r onlineSupplementalFigureS4, fig.cap=sfig_lcmm}
plot_lcmm_results(lcmm_data)
```

```{r onlineSupplementalFigureS5, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results, 'Beta-cell function')
```

```{r onlineSupplementalFigureS6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_results)
```

```{r onlineSupplementalFigureS7, fig.cap=sfig_plsda_groups_dysgly}
plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
```

```{r onlineSupplementalFigureS8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_dysgly)
```

```{r onlineSupplementalFigureS9, fig.cap=sfig_plsda_example}
plot_plsda_grouping(example_plsda_results(), 'Example groups')
```

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
output: fost::diabetologia_manuscript
---

@@

<!-- 47 characters + spaces -->
<!--Running title: NEFA and the pathogenesis of diabetes-->

<!-- First, middle initial, last -->
Author: Luke W. Johnston (1); Stewart B. Harris (2); Ravi Retnakaran
(3,4); Adria Giacca (5); Zhen Liu (1);
Richard P. Bazinet (1); and Anthony J. Hanley (1,3,6)

Institutions of origin:

@@

Word count - Abstract: 242 / 250

Word count - Main text: 3726 / 4000

@@

<!-- Reference limit 39 / 50 -->

@@

Aim: Our aim was to determine the longitudinal associations of individual
non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and
beta-cell function over 6 years in a cohort of individuals who are at-risk for
diabetes.

Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and completed an OGTT at 3 time points over
6 years. Outcome variables were calculated using the OGTT values. IS was
assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function was
assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, baseline age, alanine
aminotransferase, and family history of diabetes. NEFA were analyzed as both
concentrations (nmol/mL) and proportions (mol%) of the total fraction.

@@

## Keywords:

insulin sensitivity; beta-cell function; longitudinal cohort; non-esterified
fatty acids; diabetes pathogenesis; fatty acid composition

## Abbreviations:

- GEE: Generalized estimating equations
- LCMM: Latent curve mixed models
- PLS-DA: Partial least squares regression discriminant analysis
- IGI/IR: Insulinogenic index over HOMA-IR
- ISI: Insulin Sensitivity Index (Matsuda index)
- ISSI-2: Insulin Secretion-Sensitivity Index-2
- ALT: Alanine aminotransferase
- TAG: Triacylglyceride
- PROMISE: Prospective Metabolism and Islet Cell Evaluation Cohort
- QIC: Quasi-Likelihood Information Criterion
- MAQ: Modified Activity Questionnaire
- WC: Waist circumference

@@

```{r caption_setup}
## Include captions below using `captioner` package
fig <- captioner(prefix = 'Fig.')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner(prefix = 'ESM Fig.')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner(prefix = 'ESM Table')
cite_st <- pryr::partial(stab, display = 'cite')
# usage: cite_st('basicChar')

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
fig_gee <- fig('gee', 'Longitudinal associations of individual NEFA (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. Fully adjusted generalized estimating equation models. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')

@@

sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%. See the ESM Methods for a description of PLS-DA and an explanation of interpreting this plot.')

@@

sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). See the ESM Methods for a description of PLS-DA and an explanation of interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.')
sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA). Discriminatory ability is evident from the amount of separation between the two groups. In this case, PLS-DA was 82% accurate at correctly classifying groups. See the ESM Methods for a description of PLS-DA and for interpreting this plot.')
sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). See the ESM Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis. See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).  See the ESM Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status. See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))

@@

# Introduction

@@

# Methods

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
Participants are contacted annually by telephone. The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. Therefore, to confirm the results of the GEE
analyses in a multivariate environment (all NEFA variables in a single model),
partial least squares discriminant analysis (PLS-DA) was used to identify the
patterns of NEFA composition against beta-cell function (using ISSI-2, which had
the majority of the associations in the GEE models) and for those
who converted to or maintained dysglycemia status (either IFG, IGT, or DM) over
the 6 years. Latent class mixed models (LCMM) were used to extract the underlying
trajectories of beta-cell function (ISSI-2) over the 6 years. For more detailed
explanation of GEE, PLS-DA, and LCMM, please see the ESM Methods.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at https://figshare.com/s/a45a1e1eec5d9421eba3. Results were considered statistically significance at p<0.05,
after adjusting for multiple testing using the Benjamini-Hochberg False
Discovery Rate [@Benjamini1995a]. STROBE was used as a guideline for reporting [@Vandenbroucke2007b].

@@

```{r inlineResults}
percent.nefa <- calculate_percent_nefa_contribution(ds) 
misclass.n.percent.bcf <- calculate_plsda_misclass(plsda_results)
misclass.n.percent.dysgly <- calculate_plsda_misclass(plsda_dysgly)
percent.change.outcomes <- calculate_outcomes_pct_change(ds)
lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
dm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertDM')
predm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertPreDM')
sex.npct <- calculate_factor_npercent(ds0yr$Sex, 'Female')
ethn.npct <- calculate_factor_npercent(ds0yr$Ethnicity, 'European')
fhd.npct <- calculate_factor_npercent(ds0yr$FamHistDiab, 'Yes')
est.gee <- calculate_gee_magnitude(gee_results)
sample.outcomes <- calculate_sample(ds)
```

@@

In conclusion, we found that total NEFA was a strong predictor for lower
beta-cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lower circulating NEFA
indiscriminately to reduce the risk of diabetes are well warranted. While future
studies are needed to confirm these findings, our results reinforce the
importance of continuing to investigate the role of circulating NEFA
concentration on the natural history of diabetes.

# Acknowledgements

The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
in Family Medicine, University of Western Ontario, London, Canada for their 
expert technical assistance and dedication in their work for PROMISE.

## Funding

This study was supported by grants from the Canadian Diabetes Association (CDA),
the Canadian Institutes for Health Research, and the University of
Toronto Banting and Best Diabetes Centre; LWJ is supported by a CDA
Doctoral Student Research Award; RR is supported by a Heart and Stroke
Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
National Diabetes Management and the Ian McWhinney Chair of Family Medicine
Studies at the University of Western Ontario; RBP holds a Tier II Canada
Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
Chair in Diabetes Epidemiology.

## Duality of interest

The authors report no potential conflicts of interest relevant to this study.

## Contribution statement

The authors had the following responsibility: LWJ conducted research, analyzed
data, and wrote the paper; RR, ZL, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content. All authors read and approved the final manuscript.

Note: Values are in median (IQR), mean (SD), and n (%). The proportion of ethnic
and sex groups did not change over the 6 years. Outcome variables sample size
ranges from `r sample.outcomes` over the 3 visits.

# Figure legends

@@

# ESM Material

## ESM Methods

@@

## ESM Tables and Figures

@@

```{r ESM_Table1}
prep_qic <- gee_df %>%
    dplyr::select(
        SID, VN, lISSI2, lISI, TotalNE, Sex, Ethnicity, MET, ALT, BaseAge,
        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
        )

@@

```{r ESM_Table2}
table_distribution(ds, stab_dist)
```

```{r ESM_Fig2, fig.cap=sfig_corr}
plot_heatmap(ds)
```

```{r ESM_Fig3, fig.cap=sfig_gee}
plot_gee_results(gee_results_unadj)
```

```{r ESM_Table3_4}
table_gee(gee_results, stab_gee)
table_gee(gee_results_unadj, stab_gee_unadj)
```

```{r ESM_Fig4, fig.cap=sfig_lcmm}
plot_lcmm_results(lcmm_data)
```

```{r ESM_Fig5, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results, 'Beta-cell function')
```

```{r ESM_Fig6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_results)
```

```{r ESM_Fig7, fig.cap=sfig_plsda_groups_dysgly}
plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
```

```{r ESM_Fig8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_dysgly)
```

```{r ESM_Fig9, fig.cap=sfig_plsda_example}
plot_plsda_grouping(example_plsda_results(), 'Example groups')
```
